---
title: "Analyzing GeoMx NGS Data with GeomxTools"
author: "David Henderson, Nicole Ortogero, Zhi Yang, Rona Vitancol, Maddy Griswold"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    self_contained: false
    fig_width: 5
    fig_height: 4.5
    mainfont: Arial
    fontfamily: Arial
vignette: >
  %\VignetteIndexEntry{Analyzing GeoMx NGS Data with GeomxTools}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 5,
  fig.height = 4.5,
  dpi=200
)
```

## Introduction

Basic task when analyzing GeoMx NGS dataset outside the Digital Spatial Profiling Data Analysis (DSPDA) is to read the DCC files, PKC files, annotation file and  convert them to an object that allows user to perform their own analysis. 

GeomxTools gives users the capability to read these GeoMx output (DCC, PKC and annotation file) and convert them to NanoStringGeoMxSet, flag samples and probes that is below or above thresholds for quality control, remove samples or probes that are flagged.

### Dataset Introduction
For this demo analysis, we will use the kidney dataset that can be found here:

http://nanostring-public-share.s3-website-us-west-2.amazonaws.com/GeoScriptHub/Kidney_Dataset_for_GeomxTools.zip

Download and unzip the file. 

NanoString GeoMx DSP dataset of diabetic kidney disease (DKD) vs healthy kidney tissue. 

Seven slides were analyzed, 4 DKD and 3 healthy. Regions of Interest (ROI) were focused two different parts of a kidney’s structure: tubules or glomeruli. One glomerulus ROI contains the entirety of a glomerulus. Individual glomeruli were identified by a pathologist as either behaving relatively healthy or diseased regardless on if the tissue was DKD or healthy. Tubule ROIs were segmented into distal (PanCK) and proximal (neg) tubules. While both distal and proximal tubules are called tubules, they perform very different functions in the kidney. Segmented areas of an ROI are called Areas of Interest (AOI).


## Setting up GeomxTools and related packages

First step in using the GeomxTools package is to install the package and invoke the library command to load it into the current session.

```{r libs, message=FALSE, warning=FALSE }
# If you have not done so, install devtools to install needed R packages

# install.packages("devtools")
# devtools::install_github("Nanostring-Biostats/GeomxTools", ref = "dev")
# devtools::install_github("Nanostring-Biostats/NanoStringNCTools")

library(GeomxTools)
library(EnvStats)
library(ggiraph)
library(openxlsx)
library(lmerTest)
library(stringr)
```

## Load DCC files, PKC files and annotation file
Next step is to load the DCC, PKC and annotation file. 

- Save your PKC and annotation files in one folder, in this example, it is 
stored in kidney_demo 

- Save your DCC files as a sub-folder of the above for easy access, in this example, 
it is stored in kidney_demo/DCC_files. 

- Copy the folder path as your datadir below and change the slash to backslash. 

```{r quickstart, message=FALSE, warning=FALSE}
# Unzip and place the Kidney Dataset files downloaded from the Geoscript HUB website
# Please enter the file path address with "/" instead of "\". Say if your file is in your downloads folder, set the datadir like below.
# datadir <- file.path("C:/Users/RVitancol/Downloads/Kidney_Dataset")

datadir <- file.path("/home/rstudio/hades/Rona/GeomxTools_1/Kidney_Dataset")

# If there is none yet, make individual folders titled "dccs" "pkcs" and "annotation" 
# to build your own lab worksheet, you may need to combine the multiple lab worksheets for your experiment in to one universal document
# Please refer to the formatting and file type of the kidney example sheet
DCCFiles <- dir(file.path(datadir, "dccs"), pattern=".dcc$", full.names=TRUE, recursive=TRUE)
PKCFiles <- dir(file.path(datadir, "pkcs"), pattern=".pkc$", full.names=TRUE, recursive=TRUE)
SampleAnnotationFile <- dir(file.path(datadir, "annotation"), pattern=".xlsx$", full.names=TRUE, recursive=TRUE)

demoData <-
  suppressWarnings(readNanoStringGeoMxSet(dccFiles = DCCFiles,
                                          pkcFiles = PKCFiles,
                                          phenoDataFile = SampleAnnotationFile,
                                          phenoDataSheet = "Template",
                                          phenoDataDccColName = "Sample_ID",
                                          protocolDataColNames = c("aoi", 
                                                                   "roi"),
                                          experimentDataColNames = "panel"))
                                                                     

# Shift counts to one to mimic how DSPDA handles zero counts
demoData <- shiftCountsOne(demoData, elt="exprs", useDALogic=TRUE) 
Initial_Data <- demoData
```

### Checking the NanoStringGeoMxSet object
For users most familiar with the structure of files in DSPDA, this is how you 
can extract your initial dataset from the object as a data frame. 
You can then save it to a csv file if you need to.

```{r countmatrix, eval = TRUE}
demoData <- Initial_Data
segmentProperties <- sData(demoData)
head(segmentProperties)

# Access count matrix
CountMatrix <- exprs(demoData)

# Access Count matrix With Probe information (Probes are rows and AOIs as columns)
BioProbeCountMatrix_With_Probe_Data <- (munge(demoData, mapping = ~ exprs + `TargetName` + `ProbeID`))


# Convert the results above to wide format
BioProbeCountMatrix_With_Probe_Data <- tidyr::spread(BioProbeCountMatrix_With_Probe_Data, SampleName, exprs)
dim(BioProbeCountMatrix_With_Probe_Data)
BioProbeCountMatrix_With_Probe_Data[1:4, 1:4]
Segment_Data <- as.data.frame(sData(demoData))

# save count matrix and segment annotations in an excel

# exclude columns in sData that are list
ListCols <- sapply(sData(demoData), is.list)

Segment_Data <- sData(demoData)[!ListCols]
list_of_datasets <- list("Segment_Data" = Segment_Data, "BioProbeCountMatrix_With_Probe_Data" = BioProbeCountMatrix_With_Probe_Data)
write.xlsx(list_of_datasets, file = "Initial_Dataset.xlsx",  row.names = TRUE)

# Other useful accessor functions
head(assayData(demoData)[["exprs"]])
head(pData(demoData), 2)
protocolData(demoData)
svarLabels(demoData) 
```

## Quality Control Assessment

Users can flag samples that fail QC thresholds based on expression. This is similar to QC in DSPDA. 

Note that GeoDiff R package, a bayesian model based data analysis package is 
currently under development and planned to be released. This will include quality 
control functions based on score test, AOI QC as well as differential expression analysis. 

### Segment QC 
The setSegmentQCFlags  function will set the QC flags in the protocolData for the segment. 
The user can add a column in their annotation file titled "nuclei" for the nuclei QC.
TRUE is indicative of a QC warning for an AOI QC check that requires closer examination.

DEFAULTS for QC
minSegmentReads=1000
                                percentAligned=80,
                                      percentSaturation=50, 
                                      minNegativeCount=10, # background QC
                                      maxNTCCount=60, 
                                      minNuclei=200, #Geomx QC
                                      minArea=16000))
                                      
Cutoffs for % Stitched: 80; % Trimmed: 80 


```{r qcobject, eval = TRUE}
# Please set the QC parameters per your study, this QC runs on default parameters similar to DSPDA setting
demoData <- setSegmentQCFlags(demoData, 
                       qcCutoffs=list(minSegmentReads=1000, # Sequencing QC
                                      percentAligned=80,
                                      percentSaturation=50, 
                                      minNegativeCount=3, # background QC
                                      maxNTCCount=8704, 
                                      minNuclei=200, #Geomx QC
                                      minArea=16000)) 
```

##### Check Results
```{r checkseqqcflag,  eval = TRUE}
# TRUE means AOI did not pass the QC parameters specified
summary(sData(demoData)[["QCFlags"]])

prData <- protocolData(demoData)
QCResults <- prData[["QCFlags"]]
QCResults <- as.data.frame(QCResults)

# Append relevant variables to the the QC results dataframe. 


QCResults <- cbind(QCResults, sData(demoData)[c("Raw", "Trimmed", "Trimmed (%)", 
                                                 "DeduplicatedReads", "Saturated (%)",
                                                 "Aligned", "Aligned (%)", "Stitched",
                                                 "Stitched (%)", "NTC", "NegGeoMean", 
                                                 "area")])


# Flag sample as PASS or WARNING if sample has a QC flag
QCResults$QCStatus <- apply(prData[["QCFlags"]], 1L, function(x) {
        y <- sum(x) == 0L
        y <- ifelse(y, "PASS", "WARNING")
        return(y)
    })
```

Other columns that are available in the NanoStringGeoMxSet that can easily be added to your 
data frame can be accessed by using sData accessor function.You can add any columns here in your QC Results

```{r callsdata,  eval = TRUE}
names(sData(demoData))

# add as additional column to your QCResults, say you want to add scan name, 
# roi, aoi and slide name to your QC results
QCResults <- cbind(QCResults, sData(demoData)[c("aoi", "roi", "slide name")])
```

```{r checkQC,  eval = TRUE}
# To check how many passed and how many segments are flagged as WARNING
# View(table(QCResults$QCStatus))
# View(QCResults)

# To view an individual QC metric

# For example: NegGeoMean
# Check which samples have LowNegatives flag 

LowNegatives <- subset(demoData, select= prData[["QCFlags"]]
                      [,"LowNegatives"] == TRUE)
head(sData(LowNegatives)[c("NegGeoMean")])

# You can filter QC Results based on segments flagged as "warning" 
head(QCResults[which(QCResults$QCStatus == "WARNING"), ])

# Additionally, you can filter QC Results to check which samples have Low Negatives
head(QCResults[which(QCResults$QCStatus == "WARNING" & 
                       QCResults$LowNegatives == "TRUE"), ])

```

##### Update cutoffs for Sequencing QC
If needed, you can update the default thresholds above for sequencing cutoffs and repeat checking the QC flags steps above.

#####  Exclude samples that did not pass Sequencing QC
Samples that were flagged can be removed from analysis by subsetting. 
Subset object to exclude all that did not pass Sequencing and background QC.

```{r removeQCSampleProbe,  eval = TRUE}
QCResultsIndex <- which(apply(protocolData(demoData)[["QCFlags"]], 
                              1L, function(x) sum(x) == 0L))
QCPassed <- demoData[, QCResultsIndex]

# QC Passed is your pruned dataset without the Sequencing QC warning AOIs
# Setting demoData to contain only those that passed AOI QC
demoData <- QCPassed
dim(demoData) 
```

```{r writetools, eval = FALSE}
# Write Pruned Bioprobe QC dataset to excel
demoData <- QCPassed
# Access Count matrix With Probe information (Probes are rows and AOIs as columns)
BioProbeCountMatrix_With_Probe_Data <- (munge(demoData, mapping = ~ exprs + `TargetName` + `ProbeID`))

# Convert the results above to wide format
BioProbeCountMatrix_With_Probe_Data <- tidyr::spread(BioProbeCountMatrix_With_Probe_Data, SampleName, exprs)
dim(BioProbeCountMatrix_With_Probe_Data)

# exclude columns in sData that are list
ListCols <- sapply(sData(demoData), is.list)
Segment_Data <- sData(demoData)[!ListCols]
list_of_datasets <- list("Segment_Data" = Segment_Data, "AOI_QC_With_Probe_Data" = BioProbeCountMatrix_With_Probe_Data)
write.xlsx(list_of_datasets, file = "AOI_QC_Dataset.xlsx",  row.names = TRUE)
```


### Biological Probe QC 

Use the setBioProbeQCFlags function to set QC Flags for probes that appear to 
be outliers in the data. Set removeLocalOutliers parameter to TRUE if you want to 
remove local outliers.

Local outliers are probes that are outliers within a segment. When the removeLocalOutliers parameter is set to TRUE, the function assigns NA to the counts for the probes for the particular segment. This will then exclude it in the computation for collapsing to Target counts.

##### Set Biological Probe QC flags using default settings from DSPDA

```{r setbioprobeqcflag,  eval = TRUE}
# Generally keep the qcCutoffs parameters unchanged. Set removeLocalOutliers to 
# FALSE if you do not want to remove local outliers
demoData <- setBioProbeQCFlags(demoData, 
                               qcCutoffs=list(minProbeRatio=0.1,
                                              percentFailGrubbs=20), 
                               removeLocalOutliers=TRUE)
```

##### Check Results

```{r checkprobeQCresults}
featureData(demoData)[["QCFlags"]][1:5, 1:5]
probeQCFlags <- featureData(demoData)[["QCFlags"]]

# Check how many Grubbs local outliers were set
probeQCFlags <- probeQCFlags[, grepl("LocalGrubbsOutlier", names(probeQCFlags))]
localFlagProbes <- probeQCFlags[apply(probeQCFlags, 1, sum) > 0, ]

# How many probes are local outliers
dim(localFlagProbes)[1]

# Check probes that are flagged as Global outlier 
# TRUE means probe did not pass global probe outlier
summary(fData(demoData)[["QCFlags"]][["GlobalGrubbsOutlier"]])

# TRUE means probe did not pass low probe ratio threshold, if there is no TRUE, 
# means all probes passed low ratio threshold
summary(fData(demoData)[["QCFlags"]][["LowProbeRatio"]])

# Construct ProbeQC results dataframe
ProbeQCResults <- fData(demoData)[["QCFlags"]]
ProbeQCResults$medianCounts <- apply(assayDataElement(demoData, elt="exprs"),
                                     MARGIN=1, FUN=median)
ProbeQCResults$totalCounts <- apply(assayDataElement(demoData, elt="exprs"),
                                    MARGIN=1, FUN=sum)

#Flag sample as PASS or WARNING if sample has a QC flag
ProbeQCResults$outlierStatus <- apply(ProbeQCResults, 1L, 
                                      function(x) {
                                        y <- sum(x) == 0L
                                        y <- ifelse(y, "PASS", "WARNING")
                                        return(y)}
                                     )
# Append relevant variables to the the QC results dataframe
ProbeQCResults <- cbind(ProbeQCResults, fData(demoData)[c("ProbeRatio", "TargetName")])
dim(ProbeQCResults)

# View(ProbeQCResults)
ProbeQCResults[,!c(grepl("LocalGrubbsOutlier", names(ProbeQCResults)))]

```
##### Update cutoffs for Biological Probe QC

If you need to update the thresholds, just rerun the setBioProbeQCFlags function and set the cutoffs

##### Exclude targets that did not pass Global Biological Probe QC

```{r}  
#Subset object to exclude all that did not pass Global Bioprobe QC and Outlier Ratio QC
ProbeQCPassed <- subset(demoData, 
                     fData(demoData)[["QCFlags"]][,c("LowProbeRatio")]== FALSE &
                       fData(demoData)[["QCFlags"]][,c("GlobalGrubbsOutlier")]== FALSE)
dim(ProbeQCPassed)
demoData <- ProbeQCPassed 
```

Write Probe QC Results to excel

```{r writetools_probeqc, eval = FALSE}
# Write Pruned Bioprobe QC dataset to excel
demoData <- ProbeQCPassed 
# Access Count matrix With Probe information (Probes are rows and AOIs as columns)
BioProbeCountMatrix_With_Probe_Data <- (munge(demoData, mapping = ~ exprs + `TargetName` + `ProbeID`))

# Convert the results above to wide format
BioProbeCountMatrix_With_Probe_Data <- tidyr::spread(BioProbeCountMatrix_With_Probe_Data, SampleName, exprs)
dim(BioProbeCountMatrix_With_Probe_Data)

# exclude columns in sData that are list
ListCols <- sapply(sData(demoData), is.list)
Segment_Data <- sData(demoData)[!ListCols]
list_of_datasets <- list("Segment_Data" = Segment_Data, "Probe_QC_With_Probe_Data" = BioProbeCountMatrix_With_Probe_Data)
write.xlsx(list_of_datasets, file = "Probe_QC_Dataset.xlsx",  row.names = TRUE)
```

## Collapse to Target

##### AggregateCounts function
After cleaning the object from low counts, the counts can be collapsed to Target 
using aggregateCounts function.

```{r, eval = TRUE}
# Check how many unique targets the object has
length(unique(featureData(demoData)[["TargetName"]]))

# collapse to targets
target_demoData <- aggregateCounts(demoData)
dim(target_demoData)

# Access Count matrix With target information (Target are rows and AOIs as columns)
exprs(target_demoData)[1:5, 1:5]

```


```{r writetools_target, eval = FALSE}
# Write target dataset to excel
TargetCountMatrix <- exprs(target_demoData)

# Add annotations (scan name, roi, segment, aoi) 
TargetCountMatrix <- (munge(target_demoData, mapping = ~ exprs + `roi` + 
                              `slide name` + `segment` + `aoi`))

# convert the results above to wide format
TargetCountMatrix <- tidyr::spread(TargetCountMatrix, FeatureName, exprs)
dim(TargetCountMatrix)
TargetCountMatrix[1:4, 1:10]

# exclude columns in sData that are list
ListCols <- sapply(sData(target_demoData), is.list)
Segment_Data <- sData(target_demoData)[!ListCols]
list_of_datasets <- list("Segment_Data" = Segment_Data, "Target_Data" = t(TargetCountMatrix))
write.xlsx(list_of_datasets, file = "Target_Dataset.xlsx",  row.names = TRUE)
```


## Normalization
There is also a preloaded GeoMx DSP-DA Normalization that comes with the 
NanoStringGeoMxSet class. This includes the options to normalize on quantile, 
housekeeping or negative normalization.

```{r normalizeObject, eval = TRUE}
# Q3 norm (75th percentile) for WTA/CTA  with or without custom spike-ins
target_demoData <- normalize(target_demoData, data_type="RNA", norm_method="quant", 
                      desiredQuantile = .75, toElt = "q_norm")

# Background normalization for WTA/CTA without custom spike-in
target_demoData <- normalize(target_demoData, data_type="RNA", norm_method="neg", 
                             fromElt="exprs",  toElt="neg_norm")
# Note that Negative normalization is not acceptable for custom spike-ins in this R pipeline

assayDataElement(target_demoData, elt = "q_norm")[1:3, 1:2]
assayDataElement(target_demoData, elt = "neg_norm")[1:3, 1:2]
NormCountMatrix <- assayDataElement(target_demoData, "q_norm")

# Add annotations (scan name, roi, segment, aoi) 
# Access Count matrix With Probe information (AOIs are rows and targets as columns)
NormCountMatrix <- munge(target_demoData, mapping = ~q_norm + `roi` + 
                           `slide name` + `segment` + `aoi`)
# convert the results above to wide format
NormCountMatrix <- tidyr::spread(NormCountMatrix, FeatureName, q_norm)
dim(NormCountMatrix)
NormCountMatrix[1:4, 1:10]

```

```{r writetools_norm, eval = FALSE}
# Write target dataset to excel

# exclude columns in sData that are list
ListCols <- sapply(sData(target_demoData), is.list)
Segment_Data <- sData(target_demoData)[!ListCols]

list_of_datasets <- list("Segment_Data" = Segment_Data, "Norm_Data" = t(NormCountMatrix))
write.xlsx(list_of_datasets, file = "Norm_Dataset.xlsx",  row.names = TRUE)
```


#### Normalization Plots
To help you decide which is the best normalization to use, you can plot the 
normalization factors using heatmap option in the autoplot function

```{r normalizePlot, fig.width=4.5, fig.height=4, eval = TRUE}
# Run to generate plots to compare normalization heatmap 
p_qnorm <- autoplot(target_demoData, type = "heatmap-genes",  elt = "q_norm",
                        log2scale = FALSE, heatmapGroup = c("class", "pathology", "region")
                     )
p_qnorm

```

```{r normalizePlot2, fig.width=4.5, fig.height=4, eval = FALSE}
p_negnorm <- autoplot(target_demoData, type = "heatmap-genes",  elt = "neg_norm",
                        log2scale = FALSE, heatmapGroup = c("class", "pathology", "region")
                      )
p_negnorm

# You can use the munge function to add annotation to your data for plotting purposes
NormCountMatrix <- munge(target_demoData, mapping = ~q_norm + `roi` + 
                           `slide name` + `segment` + `aoi`)
# ggplot(NormCountMatrix, aes(x = `slide name`, y = q_norm)) + geom_boxplot() + geom_jitter()
```



## Differential Expression

Example of differential expression function you can use in your analysis.

```{r defunction, eval = FALSE}
mixedModelDE <- function( object , elt = "exprs" , modelFormula = NULL , groupVar = "group" , nCores = 1 , multiCore = TRUE , pAdjust = "BY" )
{
  if ( is.null( modelFormula ) )
  {
    modelFormula <- design( object )
  }
  mTerms <- str_trim( unlist( strsplit( unlist( attr( terms( modelFormula ) , "term.labels" ) ) , "|" , fixed = TRUE ) ) , "both" )
  if ( "1" %in% mTerms )
  {
    mTerms <- mTerms[which( !( mTerms %in% "1" ) )]
  }
  pDat <- pData( object )[,mTerms]
  for ( i in names( pDat ) )
  {
    if ( inherits( i , "character" ) )
    {
      pDat[,i] <- as.factor( pDat[,i] )
    }
  }
  if ( nCores > 1 )
  {
    require( parallel )
    deFunc <- function( i , groupVar , pDat , modelFormula , exprs )
    {
      dat <- data.frame( expr = exprs$exprs[i,] , pDat )
      lmOut <- suppressWarnings( lmerTest::lmer( modelFormula , dat ) )
      lsm <- lmerTest::ls_means( lmOut , which = groupVar , pairwise = TRUE )
      lmOut <- matrix( anova( lmOut )[groupVar,"Pr(>F)"] , ncol = 1 , dimnames = list( groupVar , "Pr(>F)" ) )
      lsmOut <- matrix( lsm[,c( "Estimate" , "Pr(>|t|)" )] , ncol = 2 , dimnames = list( gsub( groupVar , "" , rownames( lsm ) ) , c( "Estimate" , "PR(>|t|)" ) ) )
      return( list( anova = lmOut , lsmeans = lsmOut ) )
    }
    exprs <- new.env()
    exprs$exprs <- assayDataElement( object , elt = elt )
    if ( multiCore )
    {
      mixedOut <- mclapply( featureNames( object ) , deFunc , groupVar , pDat , formula( paste( "expr" , as.character( modelFormula )[2] , sep = " ~ " ) ) , exprs , mc.cores = nCores )
    }
    else
    {
      cl <- makeCluster( getOption( "cl.cores" , nCores ) )
      mixedOut <- parLapply( cl , featureNames( object ) , deFunc , groupVar , pDat , formula( paste( "expr" , as.character( modelFormula )[2] , sep = " ~ " ) ) , exprs )
      suppressWarnings( stopCluster( cl ) )
    }
    mixedOut <- rbind( array( unlist( lapply( mixedOut , function( x ) x[["anova"]] ) ) ) , array( lapply( mixedOut , function( x ) x[["lsmeans"]] ) ) )
    colnames( mixedOut ) <- featureNames( object )
    rownames( mixedOut ) <- c( "anova" , "lsmeans" )
  }
  else
  {
    deFunc <- function( expr , groupVar , pDat , modelFormula )
    {
      dat <- data.frame( expr = expr , pDat )
      lmOut <- suppressMessages( lmerTest::lmer( modelFormula , dat ) )
      lsm <- lmerTest::ls_means( lmOut , which = groupVar , pairwise = TRUE )
      lmOut <- matrix( anova( lmOut )[groupVar,"Pr(>F)"] , ncol = 1 , dimnames = list( groupVar , "Pr(>F)" ) )
      lsmOut <- matrix( lsm[,c( "Estimate" , "Pr(>|t|)" )] , ncol = 2 , dimnames = list( gsub( groupVar , "" , rownames( lsm ) ) , c( "Estimate" , "PR(>|t|)" ) ) )
      return( list( anova = lmOut , lsmeans = lsmOut ) )
    }
    mixedOut <- assayDataApply( object , 1 , deFunc , groupVar , pDat , formula( paste( "expr" , as.character( modelFormula )[2] , sep = " ~ " ) ) , elt = elt )
  }
  if ( !is.null( pAdjust ) )
  {
    mixedOut["anova",] <- p.adjust( mixedOut["anova",] , method = pAdjust )
  }
  return( mixedOut )
}
```



```{r runde, eval = FALSE}
library( lmerTest )
library( stringr )

 
design( target_demoData ) <- ~ pathology + ( 1 | slide )
pData( target_demoData )[["slide"]] <- pData( target_demoData )[["slide name"]]
pData( target_demoData )[["slide"]] <- factor( pData( target_demoData )[["slide"]] )
target_demoDataSub <- target_demoData[,which( pData( target_demoData )[["region"]] == "glomerulus" )]
target_demoDataSub <- target_demoDataSub[,which( !( pData( target_demoDataSub )[["pathology"]] %in% "NA" ) )]
target_demoDataSub <- target_demoDataSub[,which( !( pData( target_demoDataSub )[["slide"]] %in% c( "disease3" , "normal3" ) ) )]
pData( target_demoDataSub )[["slide"]] <- factor( as.character( pData( target_demoDataSub )[["slide"]] ) )
pData( target_demoDataSub )[["pathology"]] <- factor( pData( target_demoDataSub )[["pathology"]] )
mTerms <- str_trim( unlist( strsplit( gsub( "1 | " , "" , attr( terms( design( target_demoDataSub ) ) , "variables" ) , fixed = TRUE )[-1] , "+" , fixed = TRUE ) ) , "both" )
pDat <- pData( target_demoDataSub )[,mTerms]
pDat[["pathology"]] <- as.factor( pDat[["pathology"]] )

 
# options to run the DE
riSerial <- system.time(
  mixedOuts <- mixedModelDE( target_demoDataSub , elt = "q_norm" , modelFormula = NULL , groupVar = "pathology" , nCores = 1 , multiCore = TRUE , pAdjust = NULL )
)

 
riMC <- system.time(
  mixedOutmc <- mixedModelDE( target_demoDataSub , elt = "q_norm" , modelFormula = NULL , groupVar = "pathology" , nCores = 12 , multiCore = TRUE , pAdjust = NULL )
)


riP <- system.time(
  mixedOutp <- mixedModelDE( target_demoDataSub , elt = "q_norm" , modelFormula = NULL , groupVar = "pathology" , nCores = 12 , multiCore = FALSE , pAdjust = NULL )
)
```

 
```{r}
sessionInfo()
```
